Text Is All You Need

### 2023-05-04 docs/hes_final/article.qmd
spelling corrections

The GDELT Dataset has been used for a wide variety of research purposes, including for conflict prediction [@DBLP:conf/sac/ChenJY20]. While we believed this to be the best dataset available for our purposes, prior research has revealed some limitations that we must consider. GDELT’s coverage overrepresents the western world, particularly the United States and Europe. GDELT does represent Russia, India and China better than other event datasets, but it continues to underrepresent other parts of the world, particularly the global south. English is the most common language in the GDELT dataset, comprising more than 40% of the articles in its corpus [@Kwak_An_2021].

@@

::: {#fig-dgp layout-ncol=1}
```{mermaid}
%%| fig-width: 3
%%| fig-height: 5
flowchart TD
    GSG[GDELT Global Similarty Graph] -->|sample 20% of daily files| Merge{Merge on article URL,\n select relevant features}
    GKG[GDELT Global Knowledge Graph] -->|sample 20% of daily files| Merge

@@

This new dataset offered major advantages in combatting the curse of dimensionality, however, it also presented a distorted view of the world which overrepresents countries that receive more media coverage and underrepresents those that receive less, amplifying the biases already present in the GDELT dataset. Despite this limitation, the dataset was sufficient to train the encoder, which was fit to the modified dataset with the time-shifted spike metric as its target. Once training was complete, the final layer of the encoder was removed, exposing the penultimate layer of 16 dense neurons. The output of these 16 neurons would be used in the classification stage.

@@

Our target was defined differently from related work which made direct comparison less meaningful. The closest approach used a presense or absense of any armed conflict related fatality as their binary target [@10.1093/jeea/jvac025]. They focused on ROC_AUC as their metric and were able to achieve 0.83 on overall dataset and 0.75 on hard cases (sudden conflict in peaceful countries) with their text based (LDA topic extraction) model. Our best ROC_AUC score is 69% (achieved by text + country labels XGBoost/KNN model).

@@

## Ethical Considerations
This work carries significant ethical considerations, especially if it were to be operationalized. The first and most apparent is bias in the training data. All of our models showed varying performance by region, and overreliance on such models by decision makers could lead to resources being diverted away from regions that need them, but which were underrepresented in the training data and thus had less accurate results.

If models like these were to become widely used, they may be vulnerable to adversarial attacks. Large volumes of fake news content could be used to overwhelm the true signal from news articles to obscure the intentions of malicious actors, or to cause panic. Finally, and perhaps most concerningly, is the potential for conflict prediction models to deliver self-fulfilling prophesies. If policy makers become convinced of impending conflict in their own or neighboring countries because of the predictions of an AI model, they may take actions that make conflict more likely.

@@

Building an application with a user interface could be a natural extension of this work. We envision a dashboard where a user can view a world map where countries are heatmap colored according to the estimated probability of escalating conflict. Countries with highest risk can be flagged or shown to a user in a separate view. In addition to the visual part of this application, model deployment for easy inference is an important project.

>>>>

The GDELT Dataset has been used for a wide variety of research purposes, including for conflict prediction [@DBLP:conf/sac/ChenJY20]. While we believed this to be the best dataset available for our purposes, prior research has revealed some limitations that we must consider. GDELT’s coverage over-represents the western world, particularly the United States and Europe. GDELT does represent Russia, India and China better than other event datasets, but it continues to underrepresent other parts of the world, particularly the global south. English is the most common language in the GDELT dataset, comprising more than 40% of the articles in its corpus [@Kwak_An_2021].

@@

::: {#fig-dgp layout-ncol=1}
```{mermaid}
%%| fig-width: 3
%%| fig-height: 5
flowchart TD
    GSG[GDELT Global Similarity Graph] -->|sample 20% of daily files| Merge{Merge on article URL,\n select relevant features}
    GKG[GDELT Global Knowledge Graph] -->|sample 20% of daily files| Merge

@@

This new dataset offered major advantages in combatting the curse of dimensionality, however, it also presented a distorted view of the world which over-represents countries that receive more media coverage and underrepresents those that receive less, amplifying the biases already present in the GDELT dataset. Despite this limitation, the dataset was sufficient to train the encoder, which was fit to the modified dataset with the time-shifted spike metric as its target. Once training was complete, the final layer of the encoder was removed, exposing the penultimate layer of 16 dense neurons. The output of these 16 neurons would be used in the classification stage.

@@

Our target was defined differently from related work which made direct comparison less meaningful. The closest approach used a presence or absence of any armed conflict related fatality as their binary target [@10.1093/jeea/jvac025]. They focused on ROC_AUC as their metric and were able to achieve 0.83 on overall dataset and 0.75 on hard cases (sudden conflict in peaceful countries) with their text based (LDA topic extraction) model. Our best ROC_AUC score is 69% (achieved by text + country labels XGBoost/KNN model).

@@

## Ethical Considerations
This work carries significant ethical considerations, especially if it were to be operationalized. The first and most apparent is bias in the training data. All of our models showed varying performance by region, and over-reliance on such models by decision makers could lead to resources being diverted away from regions that need them, but which were underrepresented in the training data and thus had less accurate results.

If models like these were to become widely used, they may be vulnerable to adversarial attacks. Large volumes of fake news content could be used to overwhelm the true signal from news articles to obscure the intentions of malicious actors, or to cause panic. Finally, and perhaps most concerning, is the potential for conflict prediction models to deliver self-fulfilling prophesies. If policy makers become convinced of impending conflict in their own or neighboring countries because of the predictions of an AI model, they may take actions that make conflict more likely.

@@

Building an application with a user interface could be a natural extension of this work. We envision a dashboard where a user can view a world map where countries are colored according to the estimated probability of escalating conflict. Countries with highest risk can be flagged or shown to a user in a separate view. In addition to the visual part of this application, model deployment for easy inference is an important project.
